Linking watershed nutrient loading to estuary water quality with Generalized Additive Models

### 2023-08-08 Rebuttal_Letter/Rebuttal_Letter.Rmd

*My second concern about the models in eqns. 5-7 also relates to the flow and load variables. It’s stated on line 199-201 that flow-adjusted load values were used, which makes sense, but I don’t understand how this additional analysis differs from the flow-normalization in the previous section (lines 176 – 179). Why not just use the flow-adjusted loads from the previous models? Once this is clarified (or corrected), it might make more sense to change eqn 6 to include smoothers for ddate, yday, and flow-normalized load, and eqn 7 to included smoothers for ddate, yday, and load. These wouldn’t be “nested” models for AIC comparison, but would perhaps provide a better approach to evaluate loads related to natural or non-natural variation. I don’t see the value in including flow as a separate variable since it’s basically a surrogate for total load.*

**Response:** I am combining my response to both these comments since they are related. These are valid concerns and related to a comment made by Reviewer 3 (indicative that I better make a change here).

@@

*Line 77: Suggest rephrasing as “…generalized linear models where the response variable is modeled as the sum of…”*

**Response:** Suggested change made.

*Tables 1 and 2: It would be useful to indicate the approximate sampling interval for the water quality parameters. It looks like quarterly. *

**Response:** I'm out of space in the tables, but include this in the text **(addline number)**.

@@

*Figure 1: Remove “contribution” from the legend. *

**Response:** Suggested change made.

*Eqn 2: I think some additional details are needed in the text below this equation. Why was d set at 0.95? Based on the text, this seems awfully close to 1 or the “mean flow”. Also, the equation includes a lag term, t – t_j, but the size of this lag is not stated, i.e., “some historical day of observation”. Was this set explicitly by the author, and if so, is there a justification? *

**Response:** (1) I have updated the equation to (hopefully) be more clear about
the calculation. For each day in the record, the full record of flows is utilized (an admittedly not efficient calculation, but not an issue for a period of record the size of this data set). (2) I set *d* to 0.95 based on results from Robson and Dourdet (2015) and Zhang and Ball (2017), which is now discussed in **lines ...**

@@

*Line 151: It’s not clear to me how fa is a unitless term. The equations show these as simple differences of average streamflow, so wouldn’t the units be in log-cfs? *

**Response:** I worded this incorrectly. I should have phrased this "dimensionless." This is fixed in **lines ...**

@@

*Line 169: What percentage of the observations for each parameter were non-detects? I realize that setting the values as ½ the limit can introduce bias, but this may not be a big deal if there aren’t many non-detects. There does appear to be a GAM package that handles non-detects with Tobit models, but it is quite old (by software standards). It may be worth exploring. https://cran.r-project.org/web/packages/cenGAM/index.html*

**Response:** 4% of TP samples were below detection limits and 20% of NO~3~ 
samples fell below detection limits.
Early in this project 
I explored the use of the Tobit I family using the linked cenGAM package and 
unfortunately found the models result in large overestimation of concentrations.
I similarly explored the use of censored Gamma models in brms with similar 
results. Bergbusch et al. (2021) report similar findings. Conversely, Hayes 
et al. (2020) apply a censored GAM using brms without issue on microcystin 
concentration in lakes. brms appears promising for fitting censored water 
quality data, but I was frankly unable to fit reasonable models for this project
which might be due to sample size or my misunderstanding of model application.

@@

*Figure 5 caption: Consider including a reference to Table 3, as it wasn’t immediately clear to me that these models didn’t include flow or load. *

**Response:** Added a reference to Table 3.

*Table 4: As for my previous comment, I’d replace the site names with “upper”, “mid”, and “lower” and arrange them accordingly. The current arrangement is “mid”, “lower”, and “upper”. Also, it looks like several of the statistical summaries are identical, e.g. the inflow and inflow+load model for TP at TCEQ-13384. Are these simply rounding issues? And if so, are these models really all that different in terms of explained variance? *

**Respones:** I've replaced the site names and reordered the table based on location.
You are correct, the statistical summaries for several models are identical indicating that the more complicated models are not explaining additional variance.

@@

*Line 420: Replace “provide” with “prove” or “be”.*

**Response:** Fixed.

## References

Hayes, Nicole M., Heather A. Haig, Gavin L. Simpson, and Peter R. Leavitt. 2020. “Effects of Lake Warming on the Seasonal Risk of Toxic Cyanobacteria Exposure.” Limnology and Oceanography Letters 5 (6): 393–402. https://doi.org/10.1002/lol2.10164.

Robson, Barbara J., and Vincent Dourdet. 2015. “Prediction of Sediment, Particulate Nutrient and Dissolved Nutrient Concentrations in a Dry Tropical River to Provide Input to a Mechanistic Coastal Water Quality Model.” Environmental Modelling & Software 63 (January): 97–108. https://doi.org/10.1016/j.envsoft.2014.08.009.

@@

Walker, Lily M., Paul A. Montagna, Xinping Hu, and Michael S. Wetz. 2021. “Timescales and Magnitude of Water Quality Change in Three Texas Estuaries Induced by Passage of Hurricane Harvey.” Estuaries and Coasts 44 (4): 960–71. https://doi.org/10.1007/s12237-020-00846-6.

Zhang, Qian, and William P. Ball. 2017. “Improving Riverine Constituent Concentration and Flux Estimation by Accounting for Antecedent Discharge Conditions.” Journal of Hydrology 547 (April): 387–402. https://doi.org/10.1016/j.jhydrol.2016.12.052.

### 2023-08-08 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.Rmd
reorder by estuary location

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. These operating 
procedures ensure consistent collection and laboratory methods are applied 
between samples collected by different entities and under different projects. 
All sites had varying time periods and availability of water quality data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}).

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                           `NO\\textsubscript{x} (mg/L)` +
                           `TKN (mg/L)` +
                           `Chlorophyll-\\textit{a} (ug/L)` +
                           `DO (mg/L)`) ~
                           Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                         data = df, output = "data.frame") |> 
  mutate(`Station ID` = c(rep("TCEQ-13383", 5), rep("TCEQ-13384", 5),
                          rep("TCEQ-13563", 5))) |> 
  rename("Parameter" = " ") |> 
  mutate(`Station ID` = case_when(
    `Station ID` == "TCEQ-13383" ~ "TCEQ-13383, Upper-Bay",
    `Station ID` == "TCEQ-13384" ~ "TCEQ-13384, Lower-Bay",
    `Station ID` == "TCEQ-13563" ~ "TCEQ-13563, Mid-Bay"),
         Method = rep(c("EPA 365.1", "EPA 353.2", "EPA 351.2", "EPA 445.0", "ASTM D888-09(C) and TCEQ SOP V1"),3),
         `AWRL\\textsuperscript{a}`= rep(c(0.06, 0.05, 0.2, 3, NA),3),
         `Criteria\\textsuperscript{b}` = rep(c(0.21,NA,NA,11.6,5),3)) |>
  mutate(`Station ID` = fct_relevel(`Station ID`,
                                    "TCEQ-13383, Upper-Bay",
                                    "TCEQ-13563, Mid-Bay",
                                    "TCEQ-13384, Lower-Bay")) |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.") |> 
  column_spec(1, width = "2cm") |>
  column_spec(6, width = "2.5cm") |> 
  collapse_rows(columns = 1, latex_hline = "major") |> 
  footnote(alphabet = c("Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.",
                        "The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\\\\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level."),
           threeparttable = T,
           escape = F)

# modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
#                            `NO\\textsubscript{x} (mg/L)` +
#                            `TKN (mg/L)` +
#                            `Chlorophyll-\\textit{a} (ug/L)` +
#                            `DO (mg/L)`) ~
#                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
#                          data = df, output = "kableExtra", format = "latex",
#                          escape = FALSE,
#                          caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.")
```

@@

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnertQuantifyingTotalSuspended2012; @zhang_improving_2017], using Eq. \ref{eq2}:

\begin{equation}
ma(t) = \frac{\sum^{J}_{j=1}{d^{t-t_j}Q_{t_{j}}}}{\sum^{J}_{j=1}d^{t-t_j}}
\label{eq2}
\end{equation}

where *t* is the day of estimation, *d* is the discount factor (set to 0.95),
*J* is the total number of daily discharge observations, $t-t_j$ is the lag time
between the current day and some historical day of observation ($t_j$). Here,
as *d* approaches zero, the time series becomes the daily observed values and 
as *d* gets closer to one, the time series becomes the mean flow.

Flow anomaly (*fa*) is a unitless term that represents how wet or dry the 
current time period is from a previous time period [@vecchiaTrendsConcentrationsUse2009; 
@zhang_improving_2017]. Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period [Eq.\ \ref{eq3}, @zhang_improving_2017] and 
and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow [Eq.\ \ref{eq4}, @zhang_improving_2017]:

@@

# flextable::flextable(df) |> 
#   set_caption(caption = "Set of GAM models specified for each water quality parameter response.",
#               autonum = run_autonum(seq_id = "tab", bkm = "table3")) |> 
#   style(pr_t = fp_text_default(font.size = 9), part = "body") |>
#   style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
#   set_table_properties(layout = "autofit") |>
#   padding(padding.top = 0, padding.bottom = 0, part = "body")

@@

```{r fig2, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.65, fig.cap="Density plots of goodness-of-fit metrics (NSE, \\textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics.", dev='cairo_pdf'}

@@

There was no evidence of long-term changes in TP or DO concentrations at any
Lavaca Bay site (Fig.\ \ref{fig:fig5}). The upper-bay site, TCEQ-13563, had
evidence of a long-term linear increase in NO~*X*~ while chlorophyll-*a*
decreased from 2005 through 2014 (Fig.\ \ref{fig:fig5}). NO~*X*~ concentration
at the mid-bay site, TCEQ-13383, displayed an unusual periodic pattern that is
indicative of a strong influence from inflow or precipitation. The temporal GAMs
did not provide evidence of long-term trends in any of the water quality constituents
at the lower-bay TCEQ-13384 site.

```{r fig5, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines (shaded regions indicate 90\\% confidence intervals) from the temporal estuary GAM display the marginal smoothed effect of date on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.", dev='cairo_pdf'}

@@

  `TCEQ-13563` <- model1
  `TCEQ-13383` <- model2
  `TCEQ-13384` <- model3

@@

  comp <- gratia::compare_smooths(
    `TCEQ-13563`,
    `TCEQ-13383`,
    `TCEQ-13384`,
    smooths = "s(ddate)",
    data = newdata,
    overall_uncertainty = FALSE)

@@

  comp |> 
    unnest(data) |> 
    mutate(lower_ci = est + (crit * se),
           upper_ci = est - (crit * se),
           label = if_else(ddate == max(ddate),
                           as.character(model),
                           NA_character_)) |> 
    ggplot(aes(x = ddate, y = est, group = model)) +
    geom_line(aes(color = model, linetype = model), alpha = 0.75) +
    geom_ribbon(aes(ymin = lower_ci,
                    ymax = upper_ci,
                    fill = model),
                alpha = 0.25) +
    scale_x_continuous(breaks = c(2005, 2010, 2015, 2020),
                       expand = expansion(mult = c(0.05,0.05))) +
    scale_fill_manual(name = "", values = unname(palette.colors(palette = "Okabe-Ito")[2:4])) +
    scale_color_manual(name = "", values = unname(palette.colors(palette = "Okabe-Ito")[2:4])) +
    scale_linetype_manual(name = "", values = c("solid","twodash","dashed")) +
    labs(x = "Year", y = ylab,
         subtitle = subtitle) +
    theme_TWRI_print(base_family = "Arial") +
    theme(axis.title.y = element_markdown(size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          panel.grid = element_blank(),
          plot.subtitle = element_text(size = 8),
          legend.direction = "vertical",
          legend.text = element_text(size = 8),
          legend.key.width = unit(25, "points"))

@@

dplyr::bind_rows(
    tar_read(tp_13563, store = store),
    tar_read(tp_13383, store = store),
    tar_read(tp_13384, store = store),
    tar_read(no3_13563, store = store),
    tar_read(no3_13383, store = store),
    tar_read(no3_13384, store = store),
    tar_read(chla_13563, store = store),
    tar_read(chla_13383, store = store),
    tar_read(chla_13384, store = store),
    tar_read(do_13563, store = store),
    tar_read(do_13383, store = store),
    tar_read(do_13384, store = store),
    tar_read(tkn_13563, store = store),
    tar_read(tkn_13383, store = store),
    tar_read(tkn_13384, store = store)
    ) |>
  mutate(Parameter = case_when(
    Parameter == "Nitrite+Nitrate" ~ "NO\\textsubscript{x}",
    Parameter == "Chlorophyll-a" ~ "Chlorophyll-\\textit{a}",
    TRUE ~ as.character(Parameter)),
    Model = case_when(
      Model == "Temporal" ~ "Temporal",
      Model == "Flow" ~ "Inflow",
      Model == "Flow + Load" ~ "Inflow + Load",
      TRUE ~ as.character(Model)),
    Site = case_when(
      Site == "TCEQ-13563" ~ "Mid-Bay",
      Site == "TCEQ-13383" ~ "Upper-Bay",
      Site == "TCEQ-13384" ~ "Lower-Bay"
    )) |>
  mutate(Parameter = fct_relevel(Parameter, "TP", "NO\\textsubscript{x}", "Chlorophyll-\\textit{a}","TKN", "DO"),
         Model = fct_relevel(Model, "Temporal", "Inflow", "Inflow + Load"),
         Site = fct_relevel(Site, "Upper-Bay", "Mid-Bay", "Lower-Bay")) |>
  tidyr::complete(Site, Model, nesting(Parameter)) |>
  mutate(model_prob = paste0(round(AICc,1), " (", round(weight,2), ")")) |>
  mutate(model_prob = case_when(
    model_prob == "NA (NA)" ~ "-",
    model_prob != "NA (NA)" ~ model_prob
  )) |>
  select(-c(AICc, weight)) |>
  pivot_wider(names_from = Model, values_from = model_prob) |>
  select(Parameter, Site, Temporal, Inflow, `Inflow + Load`) |>
  arrange(Parameter, Site) |>
  kbl(format = "latex",
      caption = "Estuary GAM AIC\\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.",
      booktabs = TRUE, escape = FALSE) |>
  column_spec(column = 3,
              bold = c(F,F,F,F,F,F,F,F,T,T,F,T,T,F,T),
              italic = c(F,F,F,F,F,F,F,F,T,T,F,T,T,F,T)) |>
  column_spec(column = 4,
              bold = c(F,F,T,T,T,F,T,T,F,F,T,F,F,T,F),
              italic = c(F,F,T,T,T,F,T,T,F,F,T,F,F,T,F)) |>
  column_spec(column = 5,
              bold = c(T,T,F,F,F,T,F,F,F,F,F,F,F,F,F),
              italic = c(T,T,F,F,F,T,F,F,F,F,F,F,F,F,F)) |>
  collapse_rows(columns = 1, latex_hline = "major")

@@

TP and NO~3~ loadings from the Lavaca Bay watershed showed high inter-annual
variability driven primarily by fluctuations in discharge. Notably, there were
no indications of trends in flow-normalized NO~3~ and TP loads in the Navidad River.
In comparison, there was weak evidence of more recent decreases in flow-normalized NO~3~ 
(but not TP) loads in the Lavaca River watershed. While the dominant agricultural
land uses differ between the Lavaca (primarily 
grazed pasture and rangeland) and Navidad (mix of pasture and row crops) 
catchments, we did not have a reason to expect different flow normalized trends 
between the two systems from land use alone. Freshwater discharges in the 
Navidad River are regulated by the Palmetto Bend Dam forming
Lake Texana at the lower extent of the river. Lentic nitrogen uptake and 
cycling may have regulating effects that mask changes
in upstream nitrogen loadings. Additional nutrient data collection in the contributing 
tributaries of Lake Texana is needed to fully assess the role of Lake Texana
in regulating nutrient delivery to the Lavaca Bay system. However, these results
suggest that there have been no changes in the NO~3~ or TP loading from the
Navidad River system at the Lake Texana discharge point when accounting for 
variations in year to year discharge.

@@

The strong positive effect of freshwater inflow on NO~*X*~, TKN, and TP concentration
are suggestive of nonpoint watershed sources, consistent with watershed uses and
with other studies relating freshwater inflow with nutrient 
concentrations in Lavaca Bay and other estuaries [@russell_effect_2006; 
@caffreyHighNutrientPulses2007; @peierlsNonmonotonicResponsesPhytoplankton2012; 
@palmerImpactsDroughtsLow2015; @ciraPhytoplanktonDynamicsLowinflow2021]. Inflow 
had a non-linear relationship with TKN at the two upstream sites, with TKN 
increasing as freshwater inflow transitioned from low to moderate levels. 
At higher freshwater inflows, the effect was attenuated, possibly indicating a 
flushing effect at higher freshwater inflow. No relationship between TKN and 
freshwater inflow were observed at TCEQ-13384 located in the lower reach of 
Lavaca Bay. The results suggest that tidal flushing might dilute TKN and act as 
a control on the effects of freshwater inflow in the lower reaches of Lavaca Bay.
The results are also consistent with previous work that suggest processing of
organic loads in the upper Lavaca Bay or tidal portions of the Lavaca River reduce
transport of nutrients to the lower reaches of Lavaca Bay [@russell_effect_2006].

@@

Additional efforts focused on identifying relevant effect sizes, sampling 
designs, and funding mechanisms that can support long term efforts are also 
needed to adequately design such a framework. Large long-term monitoring programs 
in and around the Chesapeake Bay, San Francisco Bay, and along the Mississippi 
River have proven extremely effective at informing management actions and 
tracking progress towards long-term pollutant reduction 
goals. Similar coordinated efforts across Texas coastal watersheds would provide
useful for resource management efforts intended to protect the biological and 
water quality integrity of Texas's estuaries.

>>>>

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. These operating 
procedures ensure consistent collection and laboratory methods are applied 
between samples collected by different entities and under different projects. 
All sites had varying time periods and availability of water quality data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper-bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and the 
lower-bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}). All water quality data was collected approximately 
quarterly.

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                           `NO\\textsubscript{x} (mg/L)` +
                           `TKN (mg/L)` +
                           `Chlorophyll-\\textit{a} (ug/L)` +
                           `DO (mg/L)`) ~
                           Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                         data = df, output = "data.frame") |> 
  mutate(`Station ID` = c(rep("TCEQ-13383", 5), rep("TCEQ-13384", 5),
                          rep("TCEQ-13563", 5))) |> 
  rename("Parameter" = " ") |> 
  mutate(`Station ID` = case_when(
    `Station ID` == "TCEQ-13383" ~ "TCEQ-13383, Mid-Bay",
    `Station ID` == "TCEQ-13384" ~ "TCEQ-13384, Lower-Bay",
    `Station ID` == "TCEQ-13563" ~ "TCEQ-13563, Upper-Bay"),
         Method = rep(c("EPA 365.1", "EPA 353.2", "EPA 351.2", "EPA 445.0", "ASTM D888-09(C) and TCEQ SOP V1"),3),
         `AWRL\\textsuperscript{a}`= rep(c(0.06, 0.05, 0.2, 3, NA),3),
         `Criteria\\textsuperscript{b}` = rep(c(0.21,NA,NA,11.6,5),3)) |>
  mutate(`Station ID` = fct_relevel(`Station ID`,
                                    "TCEQ-13563, Upper-Bay",
                                    "TCEQ-13383, Mid-Bay",
                                    "TCEQ-13384, Lower-Bay")) |> 
  arrange(`Station ID`) |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.") |>
  column_spec(1, width = "2cm") |>
  column_spec(6, width = "2.5cm") |>
  collapse_rows(columns = 1, latex_hline = "major") |>
  footnote(alphabet = c("Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.",
                        "The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\\\\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level."),
           threeparttable = T,
           escape = F)
```

@@

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnertQuantifyingTotalSuspended2012; @wangSedimentConcentrationPrediction2013a; @zhang_improving_2017], using Eq. \ref{eq2}:

\begin{equation}
ma(d) = \frac{Q_j + dQ_{j-1} + \mathellipsis + d^{j-1}Q_1}{1+d+...+d^{j-1}}
\label{eq2}
\end{equation}

where *ma* at discount factor *d* is calculated using flows $(Q_1,Q_2,...,Q_j)$
for days 1 through *j*. Here, as *d* approaches zero, the time series 
becomes the daily observed values and as *d* gets closer to one, the time series 
becomes the mean flow. Although @kuhnertQuantifyingTotalSuspended2012 and 
@wangSedimentConcentrationPrediction2013a suggest using multiple covariates with
*d* set at different values from 0.75 to 0.99, @robson_prediction_2015-1 and 
@zhang_improving_2017 found substantial improvements in statistical nutrient 
models with *d*  equal to 0.95 which was adopted in this study.

Flow anomaly (*fa*) is a dimensionless term that represents how wet or dry 
(or how anomalous) the current time period is from previous time periods
[@vecchiaTrendsConcentrationsUse2009; @zhang_improving_2017]. 
Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period 
[Eq.\ \ref{eq3}, @zhang_improving_2017] and 
and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow [Eq.\ \ref{eq4}, @zhang_improving_2017]:

@@

@@

```{r fig2, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.65, fig.cap="Density plots of goodness-of-fit metrics (NSE, \\textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics. Values closer to 1 for NSE and r and values closer to 0 for PBIAS represent more ideal goodness-of-fit assessments.", dev='cairo_pdf'}

@@

There was no evidence of long-term changes in TP or DO concentrations at any
Lavaca Bay site (Fig.\ \ref{fig:fig5}). The upper-bay site, TCEQ-13563, had
evidence of a long-term linear increase in NO~*X*~ while chlorophyll-*a*
decreased from 2005 through 2014 (Fig.\ \ref{fig:fig5}). NO~*X*~ concentration
at the mid-bay site, TCEQ-13383, displayed a periodic pattern that is
indicative of a strong influence from inflow or precipitation. The temporal GAMs
did not provide evidence of long-term trends in any of the water quality constituents
at the lower-bay TCEQ-13384 site.

```{r fig5, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines (shaded regions indicate 90\\% confidence intervals) from the temporal estuary GAM (Table \\ref{tab:table3}) display the marginal smoothed effect of date on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.", dev='cairo_pdf'}

@@

  `Upper-Bay` <- model1
  `Mid-Bay` <- model2
  `Lower-Bay` <- model3

@@

  comp <- gratia::compare_smooths(
    `Upper-Bay`,
    `Mid-Bay`,
    `Lower-Bay`,
    smooths = "s(ddate)",
    data = newdata,
    overall_uncertainty = FALSE)

@@

  comp |> 
    unnest(data) |> 
    mutate(lower_ci = est + (crit * se),
           upper_ci = est - (crit * se),
           label = if_else(ddate == max(ddate),
                           as.character(model),
                           NA_character_),
           model = forcats::fct_relevel(model, "Upper-Bay", "Mid-Bay", "Lower-Bay")) |> 
    ggplot(aes(x = ddate, y = est, group = model)) +
    geom_line(aes(color = model, linetype = model), alpha = 0.75) +
    geom_ribbon(aes(ymin = lower_ci,
                    ymax = upper_ci,
                    fill = model),
                alpha = 0.25) +
    scale_x_continuous(breaks = c(2005, 2010, 2015, 2020),
                       expand = expansion(mult = c(0.05,0.05))) +
    scale_fill_manual(name = "", 
                      values = unname(palette.colors(palette = "Okabe-Ito")[2:4])) +
    scale_color_manual(name = "", 
                       values = unname(palette.colors(palette = "Okabe-Ito")[2:4])) +
    scale_linetype_manual(name = "", 
                          values = c("solid","twodash","dashed")) +
    labs(x = "Year", y = ylab,
         subtitle = subtitle) +
    theme_TWRI_print(base_family = "Arial") +
    theme(axis.title.y = element_markdown(size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          panel.grid = element_blank(),
          plot.subtitle = element_text(size = 8),
          legend.direction = "vertical",
          legend.text = element_text(size = 8),
          legend.key.width = unit(25, "points"))

@@

dplyr::bind_rows(
    tar_read(tp_13563, store = store),
    tar_read(tp_13383, store = store),
    tar_read(tp_13384, store = store),
    tar_read(no3_13563, store = store),
    tar_read(no3_13383, store = store),
    tar_read(no3_13384, store = store),
    tar_read(chla_13563, store = store),
    tar_read(chla_13383, store = store),
    tar_read(chla_13384, store = store),
    tar_read(do_13563, store = store),
    tar_read(do_13383, store = store),
    tar_read(do_13384, store = store),
    tar_read(tkn_13563, store = store),
    tar_read(tkn_13383, store = store),
    tar_read(tkn_13384, store = store)
    ) |>
  mutate(Parameter = case_when(
    Parameter == "Nitrite+Nitrate" ~ "NO\\textsubscript{x}",
    Parameter == "Chlorophyll-a" ~ "Chlorophyll-\\textit{a}",
    TRUE ~ as.character(Parameter)),
    Model = case_when(
      Model == "Temporal" ~ "Temporal",
      Model == "Flow" ~ "Inflow",
      Model == "Flow + Load" ~ "Inflow + Load",
      TRUE ~ as.character(Model)),
    Site = case_when(
      Site == "TCEQ-13563" ~ "Upper-Bay",
      Site == "TCEQ-13383" ~ "Mid-Bay",
      Site == "TCEQ-13384" ~ "Lower-Bay"
    )) |>
  mutate(Parameter = fct_relevel(Parameter, "TP", "NO\\textsubscript{x}", "Chlorophyll-\\textit{a}","TKN", "DO"),
         Model = fct_relevel(Model, "Temporal", "Inflow", "Inflow + Load"),
         Site = fct_relevel(Site, "Upper-Bay", "Mid-Bay", "Lower-Bay")) |>
  tidyr::complete(Site, Model, nesting(Parameter)) |>
  mutate(model_prob = paste0(round(AICc,1), " (", round(weight,2), ")")) |>
  mutate(model_prob = case_when(
    model_prob == "NA (NA)" ~ "-",
    model_prob != "NA (NA)" ~ model_prob
  )) |>
  select(-c(AICc, weight)) |>
  pivot_wider(names_from = Model, values_from = model_prob) |>
  select(Parameter, Site, Temporal, Inflow, `Inflow + Load`) |>
  arrange(Parameter, Site) |>
  kbl(format = "latex",
      caption = "Estuary GAM AIC\\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.",
      booktabs = TRUE, escape = FALSE) |>
  column_spec(column = 3,
              bold = c(F,F,T,F,F,F,F,F,F,T,T,T,F,T,F),
              italic = c(F,F,T,F,F,F,F,F,F,T,T,T,F,T,F)) |>
  column_spec(column = 4,
              bold = c(F,F,F,T,T,T,T,F,T,F,F,F,T,F,T),
              italic = c(F,F,F,T,T,T,T,F,T,F,F,F,T,F,T)) |>
  column_spec(column = 5,
              bold = c(T,T,F,F,F,F,F,T,F,F,F,F,F,F,F),
              italic = c(T,T,F,F,F,F,F,T,F,F,F,F,F,F,F)) |>
  collapse_rows(columns = 1, latex_hline = "major")

@@

TP and NO~3~ loadings from the Lavaca Bay watershed showed high inter-annual
variability driven primarily by fluctuations in discharge. Notably, there were
no indications of trends in flow-normalized NO~3~ and TP loads in the Navidad River.
In comparison, there was weak evidence of more recent decreases in flow-normalized NO~3~ 
(but not TP) loads in the Lavaca River watershed. While the dominant agricultural
land uses differ between the Lavaca (primarily 
grazed pasture and rangeland) and Navidad (mix of pasture and row crops) 
catchments, we did not have a reason to expect different flow normalized trends 
between the two systems from land use alone. Freshwater discharges in the 
Navidad River are regulated by the Palmetto Bend Dam forming
Lake Texana at the lower extent of the river. Lentic nitrogen uptake and 
cycling may have regulating effects that mask changes
in upstream nitrogen loadings. Additional nutrient data collection in the 
tributaries of Lake Texana is needed to fully assess the role of Lake Texana
in regulating nutrient delivery to the Lavaca Bay system. However, these results
suggest that there have been no changes in the NO~3~ or TP loading from the
Navidad River system at the Lake Texana discharge point when accounting for 
variations in year to year discharge.

@@

The strong positive effect of freshwater inflow on NO~*X*~, TKN, and TP concentration
are suggestive of nonpoint watershed sources, consistent with watershed uses and
with other studies relating freshwater inflow with nutrient 
concentrations in Lavaca Bay and other estuaries [@russell_effect_2006; 
@caffreyHighNutrientPulses2007; @peierlsNonmonotonicResponsesPhytoplankton2012; 
@palmerImpactsDroughtsLow2015; @ciraPhytoplanktonDynamicsLowinflow2021]. Inflow 
had a non-linear relationship with TKN at the two upstream sites, with TKN 
increasing as freshwater inflow transitioned from low to moderate levels. 
At higher freshwater inflows, the effect was attenuated, possibly indicating a 
flushing effect at higher freshwater inflow. No relationship between TKN and 
freshwater inflow was observed at TCEQ-13384 located in the lower reach of 
Lavaca Bay. The results suggest that tidal flushing might dilute TKN and act as 
a control on the effects of freshwater inflow in the lower reaches of Lavaca Bay.
The results are also consistent with previous work that suggest processing of
organic loads in the upper Lavaca Bay or tidal portions of the Lavaca River reduce
transport of nutrients to the lower reaches of Lavaca Bay [@russell_effect_2006].

@@

Additional efforts focused on identifying relevant effect sizes, sampling 
designs, and funding mechanisms that can support long term efforts are also 
needed to adequately design such a framework. Large long-term monitoring programs 
in and around the Chesapeake Bay, San Francisco Bay, and along the Mississippi 
River have proven extremely effective at informing management actions and 
tracking progress towards long-term pollutant reduction 
goals. Similar coordinated efforts across Texas coastal watersheds would prove
useful for resource management efforts intended to protect the biological and 
water quality integrity of Texas's estuaries.
